Self-sustained carbon monoxide oxidation oscillations on size-selected platinum nanoparticles at atmospheric pressure

### 2012-04-27 paper.tex
Incorporated Sørens corrections

A beautiful example of self-sustained oscillatory behavior in chemistry is the
colorful Belousov- Zhabotinsky reaction\cite{ZAIKIN1970} which is rather easy
to demonstrate for an audience. Such self-sustained oscillatory behavior has
been explained for systems far from equilibrium in several theoretical
works\cite{HakenBog,NicolisBog}. Similar behavior has been observed within
heterogeneous catalysis where one of the very first examples was observed by Wicke
for CO oxidation on a catalyst
of highly dispersed Pt on $\gamma$-Al$_2$O$_3$\cite{BEUSCH1972}. To gain a
deeper insight into the fundamental aspects of such oscillatory reactions this
work was followed up by Ertl on Pt(110) surfaces under well-defined conditions
\cite{EISWIRTH1986}. Ertl demonstrated how the switching between the
unreconstructed Pt(110) surface and the reconstructed 2x1-Pt(110) surface
combined with large changes of oxygen sticking coefficient on the two surfaces
could drive the self-sustained oscillations. This and many other interesting
aspects of self-sustained and forced oscillatory behavior and spatio-temporal
self-organization were investigated by Ertl and coworkers on well-defined
single crystals, see e.g. reviews\cite{IMBIHL1995,JAKUBITH1990} and in
particular the Nobel Lecture\cite{Ertl2008}. One of the reasons this has
attracted interest is the importance of such catalytic reactions demanding a
better understanding of their behavior with the goal of further improvement.

CO oxidation is one of the most studied reactions due to its importance in both
automotive catalysis and in the removal of trace CO in hydrogen
streams\cite{IbsBog,Zhang2011}. The oscillatory behavior is not, however, limited to
either Pt or CO oxidation but have been found for many other metallic catalysts
like Rh, Ir, and Pd for oxidation reactions in general, NO and CO reactions,
hydrogenation reactions \cite{Lund2000,SALES1982} and redox
systems\cite{Ibele2010}. Most of the studies
are made on single crystal or polycrystalline films or filaments mainly in
conjunction with ultra-high vacuum equipment to be able to characterize
catalyst surfaces using surface science methods. The amount of work described
on supported catalysts seems more limited due to the difficulty of
characterizing those systems and because additional complexities are present
such as mass and heat transport\cite{Meunier2008}. Reports in the literature
exist of how large
scale facilities such as an industrial ammonia fixed-bed synthesis reactor in
Germany in 1989 started oscillating for several hours. This could be modeled by
coupling of the exothermic process and the heat management in the reactor
system \cite{Morud1998} but one cannot rule out that the observed oscillations
could in some systems also be created by inherent processes on the catalyst
itself. In general such oscillations would be considered highly undesirable in
industry since they would be difficult to control and result in run-away
incidents for, in particular, exothermic reactions. This would result in
damaged catalysts due to sintering at higher temperatures. On the other hand
examples exist in literature where forced oscillations seem to improve the
overall activity\cite{IMBIHL1995,Machado2005}. One particular interesting
process here could be the methanol synthesis where a reduction of the CO$_2$
content for a shorter period of time results in a subsequent higher methanol
rate compensating for a lower methanol production while running in a
reducing H$_2$ and CO mixture alone\cite{Dynamics-Bog}. This enhancement can
be explained by an extensive wetting and un-wetting of the active Cu catalyst
dependent on the oxidation potential\cite{Hansen2002}, i.e. the active area is
strongly increased when reduced but slowly returns to the less active state
when CO$_2$ is present. This suggest that an increased activity could be
obtained by oscillating between oxidizing and reducing conditions or CO$_2$
rich/lean conditions respectively.

@@

In this work we demonstrate how a two-dimensional model system of Pt
nanoparticles on a SiO$_2$ support can be constructed that shows highly
repeatable self-sustained oscillations when initially started. The system
consists of mass selected Pt nanoparticles in the range from 2 to 10\,nm 
deposited in a microreactor with a surface area of 0.75\,cm$^2$
especially designed to achieve high sensitivity\cite{Henriksen2009}. This
method offers some advantages over the chemical synthesis methods, which also
offers the possibility of size selected nanoparticles, see e.g.
\cite{Tsang2008,Shao2011}. Particularly, surfactants or solvents
\cite{Biacchi2011}, which are required for
chemical synthesis methods, are avoided with gas-aggregation formed particles
while it is still possible to vary the coverage from sub-percent to full coverage.
It is demonstrated how the nanoparticles provides self-sustained
oscillations stable in course of several days and how various parameters can be mapped out
including the temperature dependence -- an effect that, to our knowledge, has
never been measured before and that could
not be demonstrated for a continuous film of same area in similar conditions.
By combining these results with previous studies \cite{Carlsson2006,Singh2010}
and a recent detailed study of CO oxidation on Pd using STM
\cite{Hendriksen2010} a plausible explanation for the oscillatory behavior is
presented. The low coverage and extremely efficient Pt nanoparticles illustrate
some of the pit falls that may have to be considered when making in-situ
investigations of such a system since extremely low coverages may provide the
entire conversion.

\section{Results and discussion}
An oscillation measurement is performed by mounting a sample in the setup
directly after cluster deposition and anodic bonding. Initially a light-off
CO-oxidation measurement is performed (shown in Figure~\ref{fgr:initial_treatment}) where
the temperature is increased at 3\,$^\circ$C/min until the sample ignites and
achieves full conversion of CO to CO$_2$. At a CO:O$_2$ ratio of $\sim0.08$ and a
geometrical Pt coverage of the reactor area of $\sim0.1\%$ light-off occurs at approximately
180\,$^\circ$C. The temperature ramp is typically continued up to
260\,$^\circ$C whereafter the temperature is decreased at the same rate as the
heating ramp. After the initial light-off the sample is cooled to room
temperature. When at room temperature the temperature is again increased at
3\,$^\circ$C/min until sustained oscillations occur. A typical initial
treatment is seen in Figure~\ref{fgr:initial_treatment} where both the symmetrical
light-off ramp and the subsequent increase in temperature to achieve sustained
oscillations is seen. Often the sample will show oscillations either directly
upon light-off or on the falling temperature ramp.

>>>>

A beautiful example of self-sustained oscillatory behavior in chemistry is the
colorful Belousov- Zhabotinsky reaction\cite{ZAIKIN1970} which is rather easy
to demonstrate for an audience. Such self-sustained oscillatory behavior has
been explained for systems far from equilibrium in several theoretical
works\cite{HakenBog,NicolisBog}. Similar behavior has been observed within
heterogeneous catalysis where one of the very first examples was observed by Wicke
for CO oxidation on a catalyst
of highly dispersed Pt on $\gamma$-Al$_2$O$_3$\cite{BEUSCH1972}. To gain a
deeper insight into the fundamental aspects of such oscillatory reactions this
work was followed up by Ertl and coworkers on Pt(110) surfaces under well-defined conditions
\cite{EISWIRTH1986}. Ertl demonstrated how the switching between the
unreconstructed Pt(110) surface and the reconstructed 2x1-Pt(110) surface
combined with large differences in oxygen sticking coefficient on the two surfaces
could drive the self-sustained oscillations. This and many other interesting
aspects of self-sustained and forced oscillatory behavior and spatio-temporal
self-organization were investigated by Ertl and coworkers on well-defined
single crystals, see e.g. reviews\cite{IMBIHL1995,JAKUBITH1990} and in
particular the Nobel Lecture\cite{Ertl2008}. One of the reasons this has
attracted interest is the importance of such catalytic reactions demanding a
better understanding of their behavior with the goal of further improvement.

CO oxidation is one of the most studied reactions due to its importance in both
automotive catalysis and in the removal of trace CO in hydrogen
streams\cite{IbsBog,Zhang2011}. The oscillatory behavior is not, however, limited to
either Pt or CO oxidation but have been found for many other metallic catalysts
like Rh, Ir, and Pd for oxidation reactions in general, NO and CO reactions,
hydrogenation reactions \cite{Lund2000,SALES1982} and redox
systems\cite{Ibele2010}. Most of the studies
are made on single crystal surfaces or polycrystalline films or filaments mainly in
conjunction with ultra-high vacuum equipment to be able to characterize the
catalyst surfaces using surface science methods. The amount of work described
on supported catalysts seems more limited due to the difficulty of
characterizing those systems and because additional complexities are present
such as mass and heat transport\cite{Meunier2008}. Reports in the literature
exist of how large
scale facilities such as an industrial ammonia fixed-bed synthesis reactor in
Germany in 1989 started oscillating for several hours. This could be modeled by
coupling of the exothermic process and the heat management in the reactor
system \cite{Morud1998} but one cannot rule out that the observed oscillations
could in some systems also be created by inherent processes on the catalyst
itself. In general such oscillations would be considered highly undesirable in
industry since they would be difficult to control and result in run-away
incidents for, in particular, exothermic reactions. This would result in
damaged catalysts due to sintering at higher temperatures. On the other hand
examples exist in literature where forced oscillations seem to improve the
overall activity\cite{IMBIHL1995,Machado2005}. One particular interesting
process here could be the methanol synthesis where a reduction of the CO$_2$
content for a shorter period of time results in a subsequent higher methanol
rate compensating for a lower methanol production while running in a
reducing H$_2$ and CO mixture alone\cite{Dynamics-Bog}. This enhancement can
be explained by an extensive wetting and un-wetting of the active Cu catalyst
dependent on the oxidation potential\cite{Hansen2002}, i.e. the active area is
strongly increased when reduced but slowly returns to the less active state
when CO$_2$ is present. This suggest that an increased activity could be
obtained by oscillating between oxidizing and reducing conditions or CO$_2$
rich/lean conditions respectively.

@@

In this work we demonstrate how a two-dimensional model system of Pt
nanoparticles on a SiO$_2$ support can be constructed that shows highly
repeatable self-sustained oscillations when initially started. The system
consists of mass selected Pt nanoparticles in the range from 2 to 10\,nm 
deposited in a microreactor with a surface area of 0.75\,cm$^2$
especially designed to achieve high sensitivity\cite{Henriksen2009}. This
method offers some advantages over the chemical synthesis methods, which also
offers the possibility of size selected nanoparticles, see e.g.
\cite{Tsang2008,Shao2011}. Particularly, surfactants or solvents
\cite{Biacchi2011}, which are required for
chemical synthesis methods, are avoided with gas-aggregation formed particles
while it is still possible to vary the coverage from sub-percent to full coverage.
It is demonstrated how the nanoparticles provides self-sustained
oscillations stable in course of several days and how various parameters can be mapped out
including the temperature dependence -- an effect that, to our knowledge, has
never been measured before and that could
not be demonstrated for a continuous film of same area in similar conditions.
By combining these results with previous studies \cite{Carlsson2006,Singh2010}
and a recent detailed study of CO oxidation on Pd using STM
\cite{Hendriksen2010} a plausible explanation for the oscillatory behavior is
presented. The low coverage and extremely efficient Pt nanoparticles illustrate
some of the pit falls that may have to be considered when making in-situ
investigations of such a system since a small fraction of the active sites may provide the
entire conversion.

\section{Results and discussion}
An oscillation measurement is performed by mounting a sample in the setup
directly after cluster deposition and anodic bonding. Initially a light-off
CO-oxidation measurement is performed (shown in Figure~\ref{fgr:initial_treatment}) where
the temperature is increased at 3\,$^\circ$C/min until the sample ignites and
achieves full conversion of CO to CO$_2$. At a CO:O$_2$ ratio of $\sim0.08$ and a
geometrical Pt coverage of the reactor area of $\sim0.1\%$ light-off occurs at approximately
180\,$^\circ$C. The temperature ramp is typically continued up to
260\,$^\circ$C whereafter the temperature is decreased at 3\,$^\circ$C/min.
After the initial light-off the sample is cooled to room
temperature. When at room temperature the temperature is again increased at
3\,$^\circ$C/min until sustained oscillations occur. A typical initial
treatment is seen in Figure~\ref{fgr:initial_treatment}. Often the sample will
show oscillations either directly upon light-off or on the falling temperature
ramp.
